---
title: "Assignment: Estimating Dyadic Correlations"
subtitle: "Dyadic Data Analysis"
author: "Francisco Cardozo, PhD"
date: today
---

## Overview

In this assignment you will simulate dyadic data—a child and their mother—measuring optimism about the future. You will then recover the true latent correlation using two different analytic strategies: Confirmatory Factor Analysis (CFA) and a Multilevel Model (MLM). Comparing the two approaches will deepen your understanding of when and why they yield similar or different estimates.

## Learning Objectives

- Simulate multivariate data with a known latent correlation structure
- Fit a dyadic CFA and interpret the latent factor correlation
- Fit a multilevel model to dyadic (pairwise) data and extract the intraclass correlation
- Compare CFA and MLM estimates against a known population parameter

## Instructions

### Part 1: Data Simulation (25 points)

Generate a dataset of **300 child–mother dyads**. Each individual answers **four continuous items** measuring optimism about the future. The true latent correlation between child optimism and mother optimism is **ρ = 0.70**.

Use the simulation code below as your starting point. **Replace `YOUR_STUDENT_ID` with your actual student ID** so that every student has a unique dataset.

```{r}
#| echo: true
#| eval: false
library(tidyverse)

set.seed(YOUR_STUDENT_ID)
n <- 300

# --- Latent factors (correlated at r = 0.70) ---
eta_child  <- rnorm(n)
eta_mother <- 0.70 * eta_child + sqrt(1 - 0.70^2) * rnorm(n)

# Verify the population-level correlation is close to 0.70
cor(eta_child, eta_mother)

# --- Factor loadings ---
lambda <- c(0.80, 0.75, 0.70, 0.65)

# --- Generate observed items ---
items <- map_dfc(1:4, \(j) {
    e_child  <- rnorm(n)
    e_mother <- rnorm(n)

    tibble(
        "C{j}" := lambda[j] * eta_child  + sqrt(1 - lambda[j]^2) * e_child,
        "M{j}" := lambda[j] * eta_mother + sqrt(1 - lambda[j]^2) * e_mother
    )
})

# --- Add dyad ID and total scores ---
dyad_data <- items |>
    mutate(
        dyad_id = 1:n,
        child_total  = (C1 + C2 + C3 + C4) / 4,
        mother_total = (M1 + M2 + M3 + M4) / 4,
        .before = 1
    )

dyad_data |> head(8)
```

**Deliverables:**

1. Report the sample correlation between `eta_child` and `eta_mother` (before adding measurement error)
2. Report the correlation between `child_total` and `mother_total` (after measurement error)
3. Explain in 1–2 sentences why the total-score correlation is lower than the latent correlation

### Part 2: Confirmatory Factor Analysis (35 points)

Fit a dyadic CFA in which child optimism and mother optimism are modeled as separate latent factors. You may use **lavaan or Mplus** (or both for extra credit).

#### Option A: lavaan

```{r}
#| echo: true
#| eval: false
library(lavaan)

cfa_model <- "
  # Latent factors
  child  =~ C1 + C2 + C3 + C4
  mother =~ M1 + M2 + M3 + M4

  # Latent correlation
  child ~~ mother
"

cfa_fit <- cfa(
    cfa_model,
    data = dyad_data,
    std.lv = TRUE
)

summary(cfa_fit, standardized = TRUE, fit.measures = TRUE)
```

#### Option B: Mplus

Save the data for Mplus:

```{r}
#| echo: true
#| eval: false
dyad_data |>
    select(C1:C4, M1:M4) |>
    write_delim("dyadic_optimism.dat", col_names = FALSE)
```

Then use the following input file (`dyadic_optimism_cfa.inp`):

```default
TITLE: Dyadic CFA - Child and Mother Optimism

DATA: FILE = "dyadic_optimism.dat";

VARIABLE:
  NAMES = C1 C2 C3 C4 M1 M2 M3 M4;
  USEVARIABLES = C1 C2 C3 C4 M1 M2 M3 M4;

MODEL:
  CHILD  BY C1* C2 C3 C4;
  MOTHER BY M1* M2 M3 M4;

  CHILD@1;
  MOTHER@1;

  CHILD WITH MOTHER;

OUTPUT: STDYX CINTERVAL;
```

**Deliverables:**

1. Report your model syntax
2. Report model fit indices (χ², df, p-value, CFI, TLI, RMSEA, SRMR)
3. Report the standardized factor loadings for all 8 items
4. Report the estimated latent correlation between CHILD and MOTHER with its 95% confidence interval

### Part 3: Multilevel Model (35 points)

Now estimate the dyadic association using a multilevel framework. The key idea is to restructure the data so that each row is one *person* (child or mother) nested within a *dyad*.

#### Step 1: Reshape to Long Format

```{r}
#| echo: true
#| eval: false
dyad_long <- dyad_data |>
    select(dyad_id, child_total, mother_total) |>
    pivot_longer(
        cols = c(child_total, mother_total),
        names_to = "role",
        values_to = "optimism"
    ) |>
    mutate(
        role = if_else(role == "child_total", "child", "mother"),
        role_code = if_else(role == "child", -0.5, 0.5)
    )

dyad_long |> head(8)
```

#### Step 2: Fit the Multilevel Model

Fit a model with optimism as the outcome, a fixed effect for role (to account for mean differences between children and mothers), and a random intercept for dyad:

```{r}
#| echo: true
#| eval: false
library(lmerTest)

mlm_fit <- lmer(
    optimism ~ role_code + (1 | dyad_id),
    data = dyad_long
)

summary(mlm_fit)
```

#### Step 3: Compute the ICC

Extract the variance components and compute the intraclass correlation:

```{r}
#| echo: true
#| eval: false
vc <- VarCorr(mlm_fit) |> as_tibble()

icc <- vc |>
    summarise(
        sigma2_dyad    = vcov[grp == "dyad_id"],
        sigma2_residual = vcov[grp == "Residual"],
        icc = sigma2_dyad / (sigma2_dyad + sigma2_residual)
    )

icc |>
    mutate(across(everything(), \(x) round(x, 3))) |>
    print()
```

The ICC from this model represents how much of the total variance in optimism scores is shared within dyads—this is the multilevel estimate of the dyadic correlation.

**Deliverables:**

1. Report the variance components (between-dyad and within-dyad)
2. Report the ICC value
3. Report the fixed effect for role and interpret it (do children and mothers differ in average optimism?)

### Part 4: Compare and Reflect (5 points)

Create a comparison table and answer the reflection questions.

**Comparison table:**

| Method | Estimated Correlation | True Value (ρ) | Bias |
|--------|----------------------|----------------|------|
| Total-score Pearson *r* | ? | 0.70 | ? |
| CFA latent correlation | ? | 0.70 | ? |
| MLM intraclass correlation | ? | 0.70 | ? |

*Note: Bias = Estimated − True*

**Reflection questions:**

1. Which method produced the estimate closest to the true correlation of 0.70? Why do you think that is?
2. The total-score correlation is typically *attenuated* (biased downward) compared to the latent correlation. Explain why, referencing measurement error.
3. The MLM uses total scores rather than latent variables. Would you expect the MLM-based ICC to be closer to the CFA estimate or to the Pearson *r*? Explain.
4. Under what conditions might you prefer the CFA approach over the MLM approach for estimating dyadic associations, and vice versa?

## Extra Credit (10 points)

Choose **one** of the following:

1. **Factor scores in the MLM**: Instead of total scores, extract factor scores from a single-group CFA (separate for child and mother) and use them in the multilevel model. Does this improve recovery of the true correlation?
2. **Test measurement invariance**: Constrain the factor loadings to be equal across child and mother (i.e., `(L1-L4)` labels in Mplus, or `group` in lavaan). Compare the constrained and free models using a chi-square difference test.
3. **Both approaches in both software**: Run the CFA in both lavaan and Mplus, and compare the results.

## Submission

Submit a **Word document** (`.docx`) or **PDF** containing:

1. Your R script or Quarto document with all code
2. Your Mplus input and output files (if using Mplus)
3. All tables, results, and reflection answers organized by part

## Grading Rubric

| Component | Points |
|-----------|--------|
| Part 1: Data simulation and descriptive correlations | 25 |
| Part 2: CFA — model specification, fit, and interpretation | 35 |
| Part 3: MLM — model specification, ICC, and interpretation | 35 |
| Part 4: Comparison table and reflection | 5 |
| **Total** | **100** |
| Extra credit | +10 |
