---
title: "Assignment: CFA with Missing Data"
subtitle: "Software Integration"
author: "Francisco Cardozo, PhD"
date: today
---

## Overview

In this assignment, you will simulate psychological scale data, introduce missing data for one group, and then fit a confirmatory factor analysis (CFA) model. You will compare your estimated parameters to the true values from the simulation to evaluate how well your model recovers the population parameters.

## Learning Objectives

- Practice simulating psychological data with known parameters
- Understand the impact of missing data on parameter estimation
- Gain experience running CFA models in Mplus or lavaan
- Critically evaluate model results against known true values

## Instructions

### Part 1: Data Simulation (20 points)

Create a simulated dataset with the following characteristics:

1. **Sample size**: N = 400
2. **Groups**: Two groups (e.g., "Treatment" and "Control"), with 200 participants each
3. **Latent construct**: One latent factor (e.g., "Wellbeing")
4. **Items**: 5 Likert-scale items (1-5 scale)
5. **True factor loadings**: Choose your own values between 0.5 and 0.9 for each item

**Starter code:**

```{r}
#| echo: true
#| eval: false
library(tidyverse)

set.seed(YOUR_STUDENT_ID)  # Use your student ID for reproducibility
n <- 400

# Create your simulation here
data_sim <- tibble(
    id = 1:n,
    group = rep(c("Treatment", "Control"), each = n/2),
    # Add latent trait
    theta = rnorm(n, mean = 0, sd = 1)
) %>%
    mutate(
        # Generate 5 Likert items with YOUR chosen factor loadings
        # item1 = round(pmin(5, pmax(1, 3 + λ1 * theta + rnorm(n, 0, error1)))),
        # ... continue for all 5 items
    )
```

**Deliverable**: Report your chosen factor loadings (λ₁ through λ₅) in a table.

### Part 2: Introduce Missing Data (20 points)

After creating your complete dataset, introduce missing data **only for the Treatment group**. 

Requirements:

- Make approximately 20% of the item responses missing for the Treatment group
- The Control group should have no missing data
- Missing data should be **Missing Completely at Random (MCAR)**

**Starter code:**

```{r}
#| echo: true
#| eval: false
# Function to introduce MCAR missing data
introduce_missing <- function(x, prop_missing = 0.20) {
    n <- length(x)
    missing_idx <- sample(1:n, size = round(n * prop_missing))
    x[missing_idx] <- NA
    return(x)
}

# Apply only to Treatment group
data_missing <- data_sim %>%
    mutate(
        across(
            starts_with("item"),
            ~ if_else(group == "Treatment", introduce_missing(.), .)
        )
    )

# Check missing data pattern
data_missing %>%
    group_by(group) %>%
    summarise(across(starts_with("item"), ~ mean(is.na(.))))
```

**Deliverable**: Report the percentage of missing data for each item by group.

### Part 3: CFA Model (30 points)

Fit a one-factor CFA model to your data with missing values. You may use **either Mplus or lavaan** (or both for extra credit).

#### Option A: lavaan

```{r}
#| echo: true
#| eval: false
library(lavaan)

# Select only item variables
data_cfa <- data_missing %>%
    select(starts_with("item"))

# Define model
cfa_model <- '
    Wellbeing =~ item1 + item2 + item3 + item4 + item5
'

# Fit model with FIML for missing data
fit <- cfa(cfa_model, 
           data = data_cfa, 
           std.lv = TRUE,
           missing = "fiml")  # Full Information Maximum Likelihood

summary(fit, standardized = TRUE, fit.measures = TRUE)
```

#### Option B: Mplus

```default
TITLE: CFA with Missing Data

DATA: FILE IS cfa_data.dat;

VARIABLE: 
    NAMES ARE item1 item2 item3 item4 item5;
    USEVARIABLES ARE item1-item5;
    MISSING ARE ALL (-999);  ! Or your missing value code

ANALYSIS:
    ESTIMATOR = ML;

MODEL:
    Wellbeing BY item1* item2 item3 item4 item5;
    Wellbeing@1;

OUTPUT: STANDARDIZED;
```

**Deliverable**: 

1. Report your model syntax
2. Report fit indices (χ², df, p-value, CFI, TLI, RMSEA)
3. Report unstandardized and standardized factor loadings

### Part 4: Compare to True Values (30 points)

Create a comparison table and answer the reflection questions.

**Comparison Table:**

| Item | True λ | Estimated λ | Difference | % Bias |
|------|--------|-------------|------------|--------|
| item1 | ? | ? | ? | ? |
| item2 | ? | ? | ? | ? |
| item3 | ? | ? | ? | ? |
| item4 | ? | ? | ? | ? |
| item5 | ? | ? | ? | ? |

*Note: % Bias = ((Estimated - True) / True) × 100*

**Reflection Questions:**

1. How close were your estimated factor loadings to the true values? 
2. Did the model under- or over-estimate the loadings? Is there a pattern?
3. How might the missing data in the Treatment group have affected your estimates?
4. What would you expect to happen if you increased the missing data rate to 40%?
5. Why is simulation useful for understanding statistical methods?

## Extra Credit (10 points)

Choose **one** of the following:

1. Run the CFA in **both** Mplus and lavaan and compare the results
2. Run the analysis on the **complete data** (before introducing missing values) and compare to your results with missing data
3. Try a different missing data mechanism (e.g., MAR - missing at random, where missingness depends on age or another variable) and discuss how it affects estimates

## Submission

Submit the following:

1. Your R script or Quarto document with all code
2. Your Mplus input and output files (if using Mplus)
3. A brief report (1-2 pages) with:
   - Your true parameter values
   - Missing data summary
   - Model results
   - Comparison table
   - Answers to reflection questions

## Grading Rubric

| Component | Points |
|-----------|--------|
| Data simulation with correct structure | 20 |
| Missing data correctly introduced | 20 |
| CFA model correctly specified and run | 30 |
| Comparison table and reflection | 30 |
| **Total** | **100** |
| Extra credit | +10 |
