---
title: "Software Integration"
subtitle: "Becoming Multilingual in Data Science"
author: "Francisco Cardozo, PhD"
format:
  revealjs:
    theme: simple
    logo: "../images/umlogo.png"
    footer: "EPH 752 Advanced Research Methods"
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
editor_options:
  chunk_output_type: console
---

## Agenda

1.  **The *Multilingual* Analyst**: Why data interoperability matters.
2.  **The Rosetta Stone**: Moving data between R, SPSS, SAS, and Mplus.
3.  **Comparing Mplus and R Outputs**: Understanding different outputs from the same model.
4.  **AI-Assisted Analysis**: Using AI tools responsibly and effectively.

## Why be *Multilingual*?

-   Collaborators use different tools (SPSS, SAS, Stata, Mplus).
-   Some methods are implemented better (or only) in specific software.
-   **Flexibility**: Don't get stuck because of a file format.
-   **Reproducibility**: Scripting data transformation reduces manual errors.
- Cost of software licenses.

## 1. Simulating Data in R

Instead of relying on external datasets, we will create our own. This allows us to know the **true** parameters.

```{r}
#| echo: true
library(tidyverse)

set.seed(2026)
n <- 500

# Simulate responses to a 5-item psychological scale (e.g., Anxiety)
# Underlying latent trait (theta) drives all item responses

data_sim <- tibble(
    id = 1:n,
    age = sample(18:65, n, replace = TRUE),
    group = sample(c("Male", "Female"), n, replace = TRUE),
    # Latent trait: true anxiety level (unobserved in real life)
    theta = rnorm(n, mean = 0, sd = 1)
) %>%
    mutate(
        # Generate 5 Likert items (1-5 scale) based on latent trait
        # Each item has a loading (<U+03BB>) and unique error
        item1 = round(pmin(5, pmax(1, 3 + 0.8 * theta + rnorm(n, 0, 0.5)))),
        item2 = round(pmin(5, pmax(1, 3 + 0.7 * theta + rnorm(n, 0, 0.6)))),
        item3 = round(pmin(5, pmax(1, 3 + 0.9 * theta + rnorm(n, 0, 0.4)))),
        item4 = round(pmin(5, pmax(1, 3 + 0.6 * theta + rnorm(n, 0, 0.7)))),
        item5 = round(pmin(5, pmax(1, 3 + 0.75 * theta + rnorm(n, 0, 0.55))))
    )
```

## 2. Exporting to Excel/CSV

The universal exchange format.

```{r}
#| echo: true
#| eval: false

# CSV (Comma Separated Values) - Good for interoperability
write_csv(data_sim, "data_sim.csv")

# Excel - Best for human viewing
library(writexl)
write_xlsx(data_sim, "data_sim.xlsx")
```

## 3. R $\leftrightarrow$ SPSS

Using the `haven` package to handle `.sav` files with metadata (labels).

```{r}
#| echo: true
#| eval: false
library(haven)

# Add variable labels (SPSS loves labels)
data_spss <- data_sim %>%
    mutate(
        group = as.factor(group) # SPSS handles factors as labelled integers
    )

attr(data_spss$group, "label") <- "Gender"

# Write to SPSS
write_sav(data_spss, "data_sim.sav")

# Read back
data_from_spss <- read_sav("data_sim.sav")
```

## 4. R $\leftrightarrow$ SAS

SAS `.sas7bdat` files are common in clinical trials and public health.

```{r}
#| echo: true
#| eval: false
library(haven)

# Write to SAS
write_sas(data_sim, "data_sim.sas7bdat")

# Read from SAS
data_from_sas <- read_sas("data_sim.sas7bdat")
```

## 5. R $\leftrightarrow$ Parquet

Parquet is a modern, efficient columnar format—great for big data and cross-language workflows (R, Python, Spark).

```{r}
#| echo: true
#| eval: false
library(arrow)

# Write to Parquet - fast and compressed
write_parquet(data_sim, "data_sim.parquet")

# Read from Parquet
data_from_parquet <- read_parquet("data_sim.parquet")

# Parquet preserves data types and is much faster than CSV for large files
```

**Why Parquet?**

- **Speed**: 10-100x faster than CSV for large datasets
- **Size**: Efficient compression (often 50-90% smaller)
- **Cross-platform**: Works with Python, Spark, and cloud tools

## 6. R $\leftrightarrow$ Mplus

`MplusAutomation` is the bridge. It handles the  formatting Mplus requires (no headers, specific missing values).

```{r}
#| echo: true
#| eval: false
library(MplusAutomation)

# Prepare data for Mplus
# This converts factors to numbers, handles missing values,
# and creates the .dat and .inp template
prepareMplusData(
    data_sim,
    filename = "data_sim.dat",
    inpfile = TRUE
)
```

## Challenge

1. Import a dataset you have in your computer to R. 
2. Export it to another software format. 

# Part II

## Comparing Mplus and R Outputs

**Goal**: Fit the same CFA model in both Mplus and lavaan, then compare results.

**True parameters** (from simulation):

- Factor loadings: λ₁=0.8, λ₂=0.7, λ₃=0.9, λ₄=0.6, λ₅=0.75

## Step 1: Export Data for Mplus

```{r}
#| echo: true
#| eval: false
library(MplusAutomation)

# Select only the items for CFA
data_cfa <- data_sim %>%
    select(item1, item2, item3, item4, item5)

# Export to Mplus format
prepareMplusData(
    data_cfa,
    filename = "cfa_data.dat",
    inpfile = FALSE
)
```

## Step 2: Mplus Input File

Your input file for Mplus should look like this:

```default
TITLE: CFA with 5 items - Anxiety Scale

DATA: FILE IS cfa_data.dat;

VARIABLE: 
    NAMES ARE item1 item2 item3 item4 item5;
    USEVARIABLES ARE item1-item5;

ANALYSIS:
    ESTIMATOR = ML;

MODEL:
    Anxiety BY item1* item2 item3 item4 item5;
    Anxiety@1;  ! Fix factor variance to 1

OUTPUT: STANDARDIZED MODINDICES;
```

## Step 3: Run Mplus from R

```{r}
#| echo: true
#| eval: false
library(MplusAutomation)

# Run the Mplus model
runModels("cfa_model.inp")

# Read the results back into R
mplus_results <- readModels("cfa_model.out")

# View parameter estimates
mplus_results$parameters$unstandardized
```

## Step 4: CFA in lavaan

```{r}
#| echo: true
#| eval: false
library(lavaan)

# Define the model syntax (similar to Mplus)
cfa_model <- "
    Anxiety =~ item1 + item2 + item3 + item4 + item5
"

# Fit the model
fit_lavaan <- cfa(cfa_model,
    data = data_cfa,
    std.lv = TRUE
) # Fix factor variance to 1

# View results
summary(fit_lavaan, standardized = TRUE, fit.measures = TRUE)
```

## Step 5: Extract lavaan Parameters

```{r}
#| echo: true
#| eval: false
# Get parameter estimates
parameterEstimates(fit_lavaan)

# Get standardized estimates
standardizedSolution(fit_lavaan)

# Get fit indices
fitMeasures(fit_lavaan, c("chisq", "df", "pvalue", "cfi", "tli", "rmsea"))
```

## Comparing Results

| Parameter | True Value | Mplus | lavaan |
|-----------|------------|-------|--------|
| λ₁ (item1) | 0.80 | ? | ? |
| λ₂ (item2) | 0.70 | ? | ? |
| λ₃ (item3) | 0.90 | ? | ? |
| λ₄ (item4) | 0.60 | ? | ? |
| λ₅ (item5) | 0.75 | ? | ? |

**Question**: Do both programs recover the true parameters?

## Differences to Note

| Aspect | Mplus | lavaan |
|--------|-------|--------|
| Default estimator | MLR (robust) | ML |
| First loading | Fixed to 1 | Fixed to 1 |
| Output | Separate .out file | R console |
| Syntax | `BY` | `=~` |

## Challenge

1. Do the same CFA but use the items as categorical variables.

# Part III

## AI-Assisted Data Analysis

AI tools (ChatGPT, Claude, GitHub Copilot, Cursor) are changing how we write code and conduct analyses.

How do we use AI effectively?

## What AI Can Help With

- **Code generation**: Writing R, Mplus, or Python syntax
- **Debugging**: Finding errors in your code
- **Translation**: Converting code between languages (R → Python, SPSS → R)
- **Documentation**: Explaining what code does
- **Learning**: Understanding new statistical concepts

## What AI Struggles With

- **Domain expertise**: Understanding your specific research context
- **Data interpretation**: Making substantive conclusions
- **Cutting-edge methods**: Very recent techniques not in training data
- **Verification**: AI can confidently produce incorrect code
- **Reproducibility**: Same prompt may yield different results

## Best Practices for AI in Analysis

1. **Always verify output** - Run the code, check the results
2. **Understand before using** - Don't use code you can't explain
3. **Start specific** - Detailed prompts yield better results
4. **Iterate** - Refine your prompts based on output
## Example: Good vs. Bad Prompts

:::: {.columns}

::: {.column width="50%"}
**Bad prompt:**

> "Write CFA code"

:::

::: {.column width="50%"}
**Good prompt:**

> "Write lavaan code for a one-factor CFA with 5 items (item1-item5), fix factor variance to 1, and request standardized estimates and fit indices"

:::

::::

## Example: Using AI to Translate Code

**Prompt:**

> "Convert this Mplus syntax to lavaan:
> 
> MODEL:
>     Anxiety BY item1* item2 item3 item4 item5;
>     Anxiety@1;"

**AI Output:**

```r
model <- '
    Anxiety =~ item1 + item2 + item3 + item4 + item5
'
fit <- cfa(model, data = mydata, std.lv = TRUE)
```

## Example: Using AI for Debugging

**Prompt:**

> "I'm getting this error in lavaan: 
> Error: lavaan ERROR: sample covariance matrix is not positive-definite.
> 
> My data has 5 Likert items. What could be causing this?"

AI can suggest: checking for missing data, constant variables, multicollinearity, or small sample size.

## AI Ethics in Research

- **Verification**: You are responsible for accuracy, not the AI
- **Data privacy**: Don't share sensitive data with AI tools

## AI Tools for Data Analysis

| Tool | Best For | Cost |
|------|----------|------|
| ChatGPT | General coding, explanations | Free / $20 mo |
| Claude | Longer code, nuanced tasks | Free / $20 mo |
| GitHub Copilot | IDE integration, autocomplete | $10 mo (free for students) |
| Cursor | Full IDE with AI | Free / $20 mo |

## Hands-On: AI-Assisted Coding

Try this exercise:

1. Ask an AI to write code that simulates data for a 2-factor CFA
2. Run the code in R
3. Does it work? What needed fixing?
4. Ask the AI to add missing data to the simulation

**Reflect**: What was helpful? What required your expertise?

## Summary

- AI is a **tool**, not a replacement for statistical knowledge
- **Verify everything** - AI makes confident mistakes
- Use AI to **accelerate** your work, not to bypass learning
- Your **expertise** is needed for interpretation and context

## Key Takeaways

1. **Be multilingual**: R is your hub for connecting software
2. **Simulation**: Know the truth to evaluate your methods
3. **Compare outputs**: Different software, same results (mostly)
4. **Use AI wisely**: Powerful tool with important limitations

## Questions?


