---
title: "Causal Inference"
subtitle: "Experimental and Quasi-experimental Designs"
author: "Francisco Cardozo"
format:
  revealjs:
    theme: simple
    logo: "../images/umlogo.svg"
    footer: "EPH 752 Advanced Research Methods"
    slide-number: true
    show-slide-number: all
    chalkboard: true
    link-external-newwindow: true
editor_options:
  chunk_output_type: console
---

## 1. Observational Studies & Propensity Scores ⚖️

Observational studies analyze relationships between variables without intervening in treatment assignment. 

**Propensity Score**: A statistical tool to control for selection bias and obtain more precise treatment effect estimates.

## Conceptual Stage

-   Formulate the **causal question**.
-   Use the **potential outcomes framework**.
-   "As if" it were a randomized experiment.

## Design Stage

-   Reconstruct/approximate a randomized experiment design.
-   Goal: **Independence** between treatment assignment and covariates.
-   Techniques: Matching, Stratification, Weighting (IPW).
-   **Balance Diagnosis**: Verify covariates are balanced between groups.

## Statistical Analysis Stage

-   Compare results between treatment and control groups.
-   Use appropriate statistical models (regression, etc.).
-   Adjust for propensity scores to handle selection bias.

## Example: Alcohol & Marijuana Use

> **Research Question:** What is the effect of alcohol consumption on marijuana use initiation?

Data: Selected from YRBS (Youth Risk Behavior Survey).

-   **Treatment**: Alcohol use (0/1)
-   **Outcome**: Marijuana use (0/1)
-   **Confounders**: Age, Sex, Other drug use (Cigarettes, Cocaine, Heroin, etc.)

## Hypothetical Scenarios

1.  **No Design**: Use all available data (Naive estimate).
2.  **Exclusion**: Remove outliers (e.g., polydrug users).
3.  **Stratified**: Randomize within blocks (e.g., by gender).
4.  **Propensity Score/IPW**: Reconstruct a randomized experiment using weighting.
5.  **Control for baseline**: Intervene on non-users.

## Scenario 1: Naive Estimate

```{r}
#| echo: true
#| eval: false

glm(marijuana ~ alcohol, data = data_2019, family = "binomial") %>%
    tidy(exp = T, conf.int = T)
```

*Problem*: Ignores all confounding variables.

## Scenario 4: Inverse Probability Weighting (IPW)

**Step 1: Estimate Propensity Score**

Model the probability of receiving treatment (alcohol) given covariates.

```{r}
#| echo: true
#| eval: false

# Recipe including all confounders
formula_ps <- recipe(
    alcohol ~ age + sex + q30 + q50 + q51 + q52 + q53 + q54 + q55 + q56 + q57,
    data = data_2019
)

# Logistic regression
logistic <- logistic_reg() %>% set_engine("glm")
```

## Step 2: Check Common Support

Ensure there is overlap in the probability of treatment between groups.

```{r}
#| echo: true
#| eval: false

# Histogram of propensity scores by group
to_ipw %>%
    ggplot(aes(.pred_Yes, fill = alcohol)) +
    geom_histogram(position = position_dodge(), binwidth = 0.05)
```

## Step 3: Calculate Weights

$$w = \frac{T}{e(X)} + \frac{1-T}{1-e(X)}$$

```{r}
#| echo: true
#| eval: false
data_ipw <- to_ipw |>
    mutate(
        ipw = case_when(
            alcohol == "Yes" ~ 1 / .pred_Yes,
            alcohol == "No" ~ 1 / (1 - .pred_Yes)
        ),
        ipw = importance_weights(ipw)
    )
```

## Step 4: Check Balance

Verify that after weighting, the groups look similar on covariates.

```{r}
#| echo: true
#| eval: false
# Run diagnostics on weighted data
# ...
```

## Step 5: Estimate Causal Effect

Run the outcome model weighted by the IPW.

```{r}
#| echo: true
#| eval: false

fit(causal_effect_wflow, data = finaldata) |>
    tidy(exp = TRUE, conf.int = TRUE)
```

## 2. Difference in Differences (DiD) 🔀

-   Evaluates average change after an intervention implementation.
-   Estimates the **counterfactual** based on a comparison group trend.
-   **Parallel Trends Assumption**: Without intervention, the treatment and control groups would have followed parallel paths.

## Example: Marijuana Legalization

> **Question:** What is the effect of marijuana legalization (California, 2016-2018) on adolescent marijuana use?

-   **Treatment Group**: California (CA)
-   **Control Group**: Florida (FL)
-   **Pre-Period**: 2017 (Early implementation)
-   **Post-Period**: 2019 (Full sales)

## DiD Visualization

```{r}
#| echo: true
#| eval: false

diff_diff %>%
    ggplot(aes(x = year, y = nmarijuana, color = state, group = state)) +
    geom_smooth(method = "glm", se = F) +
    labs(y = "Marijuana Prevalence", x = "Year")
```

## Estimating DiD

Model:
$$Y = \beta_0 + \beta_1 \text{Treatment} + \beta_2 \text{Time} + \beta_3 (\text{Treatment} \times \text{Time}) + \epsilon$$

$\beta_3$ is the **Difference-in-Differences estimator**.

```{r}
#| echo: true
#| eval: false

formula_did <- recipe(
    marijuana ~ year + treatment + interaction,
    data = diff_diff2
)

fit(did_wflow, data = diff_diff2) |>
    tidy(conf.int = TRUE, exp = T)
```

## Summary 🥳

-   **Observational Studies**: Require careful design to mimic experiments.
-   **Propensity Scores**: Help balance groups (Matching, IPW).
-   **Difference in Differences**: Controls for time-invariant unobserved confounders by comparing trends.
