---
title: "Causal Inference"
subtitle: "From Association to Causation"
author: "Francisco Cardozo, PhD"
brand: false
engine: knitr
format:
  revealjs:
    theme: [simple, slides-styles.scss]
    logo: "../images/umlogo.png"
    footer: "EPH 752 Advanced Research Methods"
    chalkboard: true
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
    code-overflow: wrap
    scrollable: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| include: false

library(tidyverse)
library(broom)
library(knitr)

set.seed(2026)
```

# Part I: The Causal Question {background-color="#005030"}

## Why Causation Matters {.smaller}

:::: columns
::: {.column width="50%"}

Most research questions in public health are <mark>causal</mark>:

- Does smoking **cause** lung cancer?
- Does a prevention program **reduce** substance use?
- Does Medicaid expansion **improve** health outcomes?

Yet most of our data is <mark>observational</mark>. We see associations, but associations can be misleading.

:::

::: {.column width="50%"}

::: {.callout-warning}
## The Fundamental Problem

**Association $\neq$ Causation**

Ice cream sales and drowning deaths are correlated. This does not mean ice cream causes drowning — both are driven by summer heat (a confounder).

The challenge of causal inference is extracting the **causal signal** from observational data where confounders lurk.
:::

:::
::::

## The Potential Outcomes Framework {.smaller}

:::: columns
::: {.column width="50%"}

The <mark>Rubin Causal Model</mark> defines causation through **potential outcomes**:

For each individual $i$:

- $Y_i(1)$: the outcome if **treated**
- $Y_i(0)$: the outcome if **not treated**

The <mark>individual causal effect</mark> is:

$$\tau_i = Y_i(1) - Y_i(0)$$

:::

::: {.column width="50%"}

::: {.callout-important}
## The Fundamental Problem of Causal Inference

We can **never** observe both $Y_i(1)$ and $Y_i(0)$ for the same individual. Each person is either treated or not — we only see **one** potential outcome.

| Person | $Y(1)$ | $Y(0)$ | $\tau$ |
|--------|--------|--------|--------|
| Alice  | 8      | ?      | ?      |
| Bob    | ?      | 5      | ?      |
| Carol  | 7      | ?      | ?      |
| Dave   | ?      | 4      | ?      |

The missing values are the **counterfactual** — what *would have happened* under the other condition.
:::

:::
::::

## Average Treatment Effects {.smaller}

:::: columns
::: {.column width="50%"}

Since individual effects are unobservable, we focus on <mark>average</mark> effects:

**Average Treatment Effect (ATE)**:

$$\text{ATE} = E[Y(1) - Y(0)] = E[Y(1)] - E[Y(0)]$$

The average causal effect across the entire population.

**Average Treatment Effect on the Treated (ATT)**:

$$\text{ATT} = E[Y(1) - Y(0) \mid T = 1]$$

The average effect among those who actually received treatment.

:::

::: {.column width="50%"}

::: {.callout-tip}
## Why the Distinction Matters

ATE and ATT are the <mark>same</mark> when treatment is randomly assigned (because treated and untreated groups are exchangeable).

They <mark>differ</mark> in observational studies when the people who select into treatment are systematically different from those who don't.

**Example**: People who choose to exercise (treated) may be healthier at baseline. The effect of exercise *among exercisers* (ATT) may differ from the effect *if everyone exercised* (ATE).
:::
:::
::::

## Randomization: The Gold Standard {.smaller}

:::: columns
::: {.column width="50%"}

In a <mark>randomized experiment</mark>, treatment assignment is independent of potential outcomes:

$$T \perp\!\!\!\perp \{Y(1), Y(0)\}$$

This means:

- $E[Y(1) \mid T=1] = E[Y(1)]$
- $E[Y(0) \mid T=0] = E[Y(0)]$

So the <mark>simple difference in means</mark> identifies the ATE:

$$\hat{\text{ATE}} = \bar{Y}_{T=1} - \bar{Y}_{T=0}$$

:::

::: {.column width="50%"}

Randomization works because it creates <mark>balance</mark> on all covariates — both observed and unobserved — between treatment and control groups.

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4

set.seed(2026)
n <- 500
rct_data <- tibble(
    age = rnorm(n, 40, 10),
    treatment = sample(c(0, 1), n, replace = TRUE)
) |>
    mutate(treatment_label = ifelse(treatment == 1, "Treated", "Control"))

ggplot(rct_data, aes(x = age, fill = treatment_label)) +
    geom_density(alpha = 0.5) +
    scale_fill_manual(
        values = c("Control" = "#005030", "Treated" = "#F47321")
    ) +
    labs(
        x = "Age",
        y = "Density",
        fill = "Group",
        title = "Randomization Balances Covariates"
    ) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "top")
```

:::
::::

## Simulation: Why Randomization Works {.smaller}

Let's simulate a world where we know the truth. A prevention program reduces risk scores by <mark>5 points</mark>, but sicker people are more likely to seek treatment:

```{r}
#| echo: true

set.seed(2026)
n <- 2000
true_effect <- -5

sim_data <- tibble(
    baseline_risk = rnorm(n, 50, 10),
    treatment_rct = sample(c(0, 1), n, replace = TRUE),
    treatment_obs = rbinom(n, 1, plogis(-2 + 0.06 * baseline_risk)),
    y_rct = baseline_risk + true_effect * treatment_rct + rnorm(n, 0, 5),
    y_obs = baseline_risk + true_effect * treatment_obs + rnorm(n, 0, 5)
)
```

- `treatment_rct`: randomly assigned (coin flip)
- `treatment_obs`: assigned based on `baseline_risk` (sicker people more likely to be treated)

## What Happens When We Estimate? {.smaller}

:::: columns
::: {.column width="50%"}

<mark>Randomized experiment</mark>:

```{r}
#| echo: true

lm(y_rct ~ treatment_rct, data = sim_data) |>
    tidy(conf.int = TRUE) |>
    filter(term == "treatment_rct") |>
    select(term, estimate, std.error, conf.low, conf.high) |>
    mutate(across(where(is.numeric), \(x) round(x, 2))) |>
    kable()
```

The estimate is close to the true effect of $-5$.

:::

::: {.column width="50%"}

<mark>Observational study</mark> (naive estimate):

```{r}
#| echo: true

lm(y_obs ~ treatment_obs, data = sim_data) |>
    tidy(conf.int = TRUE) |>
    filter(term == "treatment_obs") |>
    select(term, estimate, std.error, conf.low, conf.high) |>
    mutate(across(where(is.numeric), \(x) round(x, 2))) |>
    kable()
```

The estimate is <mark>biased</mark> — it may even have the wrong sign! Sicker people select into treatment, confounding the effect.

:::
::::

. . .

> The same treatment, the same effect, but <mark>confounding</mark> from selection bias distorts the observational estimate. This is why we need causal inference methods.

# Part II: Observational Studies & Selection Bias {background-color="#F47321"}

## Rubin's Three Stages {.smaller}

:::: columns
::: {.column width="50%"}

Donald Rubin's framework for causal inference from observational data involves three sequential stages:

**1. Conceptual Stage**

- Formulate the <mark>causal question</mark>
- Define treatment, outcome, and the potential outcomes
- Think "as if" designing a randomized experiment

**2. Design Stage**

- Reconstruct a <mark>quasi-randomized</mark> comparison
- Goal: make treatment assignment *as if* independent of covariates
- Techniques: matching, stratification, weighting

:::

::: {.column width="50%"}

**3. Analysis Stage**

- Compare outcomes between treatment and control groups
- Use statistical models adjusted for the design
- Conduct sensitivity analyses

::: {.callout-important}
## Key Principle

The <mark>design</mark> stage should be done **without looking at outcomes**. Like designing an RCT before collecting data, the goal is to construct balanced groups **before** estimating effects. This prevents researchers from fishing for designs that yield desired results.
:::
:::
::::

## Selection Bias: A Closer Look {.smaller}

:::: columns
::: {.column width="50%"}

<mark>Selection bias</mark> occurs when the treated and untreated groups differ systematically on characteristics that also affect the outcome.

$$E[Y \mid T=1] - E[Y \mid T=0] = \underbrace{\text{ATE}}_{\text{causal}} + \underbrace{\text{Bias}}_{\text{selection}}$$

The naive difference in means conflates the treatment effect with pre-existing differences between groups.

:::

::: {.column width="50%"}

Let's see this with our simulated data:

```{r}
#| echo: true

sim_data |>
    summarise(
        mean_risk = round(mean(baseline_risk), 1),
        n = n(),
        .by = treatment_obs
    ) |>
    mutate(group = ifelse(treatment_obs == 1, "Treated", "Control")) |>
    select(group, mean_risk, n) |>
    kable()
```

The treated group has a <mark>higher</mark> baseline risk (they were sicker to begin with). Any naive comparison is contaminated by this imbalance.

:::
::::

## Key Assumptions for Causal Inference {.smaller}

To identify causal effects from observational data, we need three assumptions:

:::: columns
::: {.column width="33%"}

::: {.callout-note}
## 1. Ignorability (No Unmeasured Confounding)

$$T \perp\!\!\!\perp \{Y(1), Y(0)\} \mid X$$

Conditional on observed covariates $X$, treatment assignment is "as if" random. This is the <mark>strongest</mark> and most untestable assumption.
:::

:::

::: {.column width="33%"}

::: {.callout-note}
## 2. Positivity (Overlap)

$$0 < P(T=1 \mid X=x) < 1$$

For every combination of covariate values, there is a <mark>nonzero probability</mark> of receiving either treatment or control. No subgroup is deterministically assigned.
:::

:::

::: {.column width="34%"}

::: {.callout-note}
## 3. SUTVA (Stable Unit Treatment Value)

Each individual's potential outcomes are <mark>not affected</mark> by other individuals' treatment assignments (no interference), and there is only <mark>one version</mark> of each treatment level.
:::

:::
::::

. . .

> If these assumptions hold, we can use propensity score methods to estimate causal effects. If they don't — especially ignorability — our estimates may still be biased.

# Part III: Propensity Score Methods {background-color="#005030"}

## What Is a Propensity Score? {.smaller}

:::: columns
::: {.column width="50%"}

The <mark>propensity score</mark> is the probability of receiving treatment given observed covariates:

$$e(X_i) = P(T_i = 1 \mid X_i)$$

Rosenbaum and Rubin (1983) showed that if ignorability holds given $X$, it also holds given $e(X)$:

$$T \perp\!\!\!\perp \{Y(1), Y(0)\} \mid e(X)$$

This is powerful: instead of balancing on many covariates simultaneously, we can balance on a <mark>single scalar</mark>.

:::

::: {.column width="50%"}

::: {.callout-tip}
## Intuition

The propensity score summarizes all measured confounders into one number. Two individuals with the same propensity score, one treated and one untreated, are like a mini randomized comparison.

Think of it as creating an "apples-to-apples" comparison: matching a treated person who had a 70% chance of being treated with an untreated person who *also* had a 70% chance.
:::

**Common uses of propensity scores:**

| Method | How PS is used |
|--------|---------------|
| **Matching** | Pair treated and untreated with similar $e(X)$ |
| **Stratification** | Group by PS quintiles, compare within strata |
| **Weighting (IPW)** | Weight observations by inverse of PS |
| **Regression adjustment** | Include PS as a covariate |

:::
::::

## Step 1: Estimate the Propensity Score {.smaller}

We estimate $e(X)$ with a logistic regression of treatment on confounders:

```{r}
#| echo: true

ps_model <- glm(
    treatment_obs ~ baseline_risk,
    data = sim_data,
    family = binomial
)

sim_data <- sim_data |>
    mutate(ps = predict(ps_model, type = "response"))

tidy(ps_model) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    kable()
```

. . .

Higher baseline risk $\to$ higher probability of treatment, confirming the selection mechanism.

## Step 2: Check Common Support (Overlap) {.smaller}

Before using propensity scores, we must verify that both groups have <mark>overlapping</mark> distributions:

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4.5

sim_data |>
    mutate(group = ifelse(treatment_obs == 1, "Treated", "Control")) |>
    ggplot(aes(x = ps, fill = group)) +
    geom_histogram(
        position = position_dodge(),
        bins = 40,
        alpha = 0.7
    ) +
    scale_fill_manual(
        values = c("Control" = "#005030", "Treated" = "#F47321")
    ) +
    labs(
        x = "Propensity Score",
        y = "Count",
        fill = "Group",
        title = "Propensity Score Distributions by Treatment Group"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "top")
```

Good overlap means for any propensity score value, we have both treated and untreated individuals to compare.

## Step 3: Inverse Probability Weighting (IPW) {.smaller}

:::: columns
::: {.column width="50%"}

IPW creates a <mark>pseudo-population</mark> where treatment is independent of confounders by weighting each observation by the inverse of their probability of receiving the treatment they actually received.

**Intuition**: A treated person with a *low* propensity score is "surprising" — they received treatment despite having characteristics that usually lead to no treatment. They get a <mark>large weight</mark> because they represent many similar people in the population who would not have been treated.

:::

::: {.column width="50%"}

::: {.callout-note}
## Why This Works

In the weighted pseudo-population:

- Treated individuals who "look like" controls get upweighted
- Control individuals who "look like" treated get upweighted

This breaks the association between covariates and treatment, mimicking randomization.
:::

| Person | Treatment | $e(X)$ | Weight |
|--------|-----------|--------|--------|
| A      | 1         | 0.8    | $1/0.8 = 1.25$ |
| B      | 1         | 0.2    | $1/0.2 = 5.00$ |
| C      | 0         | 0.8    | $1/0.2 = 5.00$ |
| D      | 0         | 0.2    | $1/0.8 = 1.25$ |

:::
::::

## Step 3: Computing IPW Weights {.smaller}

```{r}
#| echo: true

sim_data <- sim_data |>
    mutate(
        ipw = case_when(
            treatment_obs == 1 ~ 1 / ps,
            treatment_obs == 0 ~ 1 / (1 - ps)
        )
    )

sim_data |>
    summarise(
        min_weight = round(min(ipw), 2),
        median_weight = round(median(ipw), 2),
        max_weight = round(max(ipw), 2),
        mean_weight = round(mean(ipw), 2),
        .by = treatment_obs
    ) |>
    mutate(group = ifelse(treatment_obs == 1, "Treated", "Control")) |>
    select(group, min_weight, median_weight, max_weight, mean_weight) |>
    kable()
```

. . .

::: {.callout-warning}
## Watch for Extreme Weights

Very large weights (from propensity scores near 0 or 1) can create instability. Common remedies include <mark>trimming</mark> (dropping extreme PS values) or <mark>stabilized weights</mark> that multiply by $P(T)$ or $P(1-T)$.
:::

## Step 4: Check Balance After Weighting {.smaller}

The purpose of IPW is to create balance. Let's verify:

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4.5

balance_data <- bind_rows(
    sim_data |>
        mutate(
            group = ifelse(treatment_obs == 1, "Treated", "Control"),
            method = "Unweighted"
        ),
    sim_data |>
        mutate(
            group = ifelse(treatment_obs == 1, "Treated", "Control"),
            method = "IPW-Weighted"
        )
)

unweighted_means <- sim_data |>
    summarise(mean_risk = mean(baseline_risk), .by = treatment_obs)

weighted_means <- sim_data |>
    summarise(
        mean_risk = weighted.mean(baseline_risk, w = ipw),
        .by = treatment_obs
    )

bind_rows(
    unweighted_means |> mutate(method = "Unweighted"),
    weighted_means |> mutate(method = "IPW-Weighted")
) |>
    mutate(
        group = ifelse(treatment_obs == 1, "Treated", "Control"),
        mean_risk = round(mean_risk, 2)
    ) |>
    select(method, group, mean_risk) |>
    pivot_wider(names_from = group, values_from = mean_risk) |>
    mutate(Difference = Treated - Control) |>
    kable()
```

. . .

After weighting, the groups are <mark>balanced</mark> on baseline risk (the confounding is removed).

## Step 5: Estimate the Causal Effect {.smaller}

:::: columns
::: {.column width="50%"}

Now we estimate the treatment effect using the IPW-weighted regression:

```{r}
#| echo: true

ipw_model <- lm(
    y_obs ~ treatment_obs,
    data = sim_data,
    weights = ipw
)

bind_rows(
    tidy(lm(y_obs ~ treatment_obs, data = sim_data)) |>
        filter(term == "treatment_obs") |>
        mutate(method = "Naive (unadjusted)"),
    tidy(ipw_model) |>
        filter(term == "treatment_obs") |>
        mutate(method = "IPW-weighted")
) |>
    select(method, estimate, std.error) |>
    mutate(
        across(where(is.numeric), \(x) round(x, 2)),
        true_effect = true_effect
    ) |>
    kable()
```

:::

::: {.column width="50%"}

::: {.callout-tip}
## Reading the Results

- **Naive estimate**: biased because it doesn't account for selection
- **IPW estimate**: much closer to the true effect of $-5$
- The IPW successfully removed the confounding from baseline risk
:::

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4

tibble(
    Method = c("Naive", "IPW", "Truth"),
    Estimate = c(
        coef(lm(y_obs ~ treatment_obs, data = sim_data))["treatment_obs"],
        coef(ipw_model)["treatment_obs"],
        true_effect
    )
) |>
    mutate(Method = factor(Method, levels = c("Naive", "IPW", "Truth"))) |>
    ggplot(aes(x = Method, y = Estimate, fill = Method)) +
    geom_col(alpha = 0.8, width = 0.6) +
    geom_hline(
        yintercept = true_effect,
        linetype = "dashed",
        color = "red",
        linewidth = 1
    ) +
    scale_fill_manual(
        values = c("Naive" = "grey60", "IPW" = "#005030", "Truth" = "#F47321")
    ) +
    labs(
        y = "Estimated Treatment Effect",
        title = "Bias Correction via IPW"
    ) +
    theme_minimal(base_size = 13) +
    theme(legend.position = "none")
```

:::
::::

## The IPW Workflow: Summary {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 2

tibble(
    step = factor(
        c(
            "1. Estimate\nPropensity Score",
            "2. Check\nOverlap",
            "3. Compute\nIPW Weights",
            "4. Check\nBalance",
            "5. Estimate\nCausal Effect"
        ),
        levels = c(
            "1. Estimate\nPropensity Score",
            "2. Check\nOverlap",
            "3. Compute\nIPW Weights",
            "4. Check\nBalance",
            "5. Estimate\nCausal Effect"
        )
    ),
    x = 1:5,
    y = 1
) |>
    ggplot(aes(x = x, y = y)) +
    geom_tile(width = 0.85, height = 0.6, fill = "#005030", alpha = 0.85) +
    geom_text(aes(label = step), color = "white", size = 4, fontface = "bold") +
    geom_segment(
        data = tibble(x = 1.5:4.5, xend = 1.5:4.5, y = 1, yend = 1),
        aes(x = x - 0.07, xend = x + 0.07, y = y, yend = y),
        arrow = arrow(length = unit(0.15, "cm")),
        color = "#F47321",
        linewidth = 1.5
    ) +
    xlim(0.3, 5.7) +
    ylim(0.4, 1.6) +
    theme_void()
```

| Step | Action | Key Check |
|------|--------|-----------|
| 1. Propensity Score | Logistic regression of $T$ on $X$ | Model includes all confounders |
| 2. Overlap | Histogram of PS by group | Both groups span similar PS range |
| 3. Weights | $w = T/e(X) + (1-T)/(1-e(X))$ | No extreme weights |
| 4. Balance | Compare weighted covariate means | Standardized differences < 0.1 |
| 5. Outcome | Weighted regression of $Y$ on $T$ | Interpret as ATE |

::: {.callout-important}
## The Critical Assumption

All of this rests on <mark>no unmeasured confounding</mark>. If important confounders are missing from the propensity score model, the estimate remains biased. Always conduct **sensitivity analyses** to assess robustness.
:::

## Other Propensity Score Methods {.smaller}

:::: columns
::: {.column width="50%"}

**Propensity Score Matching**

- Pair each treated individual with one or more untreated individuals with similar propensity scores
- Most common: 1:1 nearest-neighbor matching
- Estimates the <mark>ATT</mark> (effect on the treated)
- Can discard unmatched controls, reducing sample size

**Stratification (Subclassification)**

- Divide the sample into strata (typically 5) based on PS quintiles
- Estimate treatment effect within each stratum
- Average across strata for an overall estimate
- Cochran (1968): 5 strata remove ~90% of confounding bias

:::

::: {.column width="50%"}

**Doubly Robust Estimation**

- Combines propensity score weighting **and** outcome regression
- Consistent if <mark>either</mark> the PS model or the outcome model is correctly specified
- Provides an extra layer of protection against model misspecification

::: {.callout-tip}
## Practical Recommendation

In practice, <mark>IPW</mark> is the most common starting point. If you suspect model misspecification, consider **doubly robust** methods. Use **matching** when you want interpretable 1:1 comparisons and the ATT is your target.

R packages: `MatchIt`, `WeightIt`, `cobalt`, `CBPS`
:::

:::
::::

# Part IV: Difference-in-Differences {background-color="#F47321"}

## The Idea Behind DiD {.smaller}

:::: columns
::: {.column width="50%"}

<mark>Difference-in-Differences (DiD)</mark> exploits a **natural experiment**: a policy or event that affects one group but not another.

Instead of comparing treated vs. untreated at one point, we compare <mark>changes over time</mark>:

$$\text{DiD} = \underbrace{(\bar{Y}_{T,\text{post}} - \bar{Y}_{T,\text{pre}})}_{\text{change in treated}} - \underbrace{(\bar{Y}_{C,\text{post}} - \bar{Y}_{C,\text{pre}})}_{\text{change in control}}$$

The control group's change estimates what <mark>would have happened</mark> to the treated group without the intervention.

:::

::: {.column width="50%"}

::: {.callout-important}
## The Parallel Trends Assumption

DiD requires that, in the absence of treatment, both groups would have followed <mark>parallel trajectories</mark>.

We cannot test this directly (it's about a counterfactual), but we can check whether trends were parallel **before** the intervention.

If pre-treatment trends diverge, DiD estimates are biased.
:::

:::
::::

## The DiD Estimator {.smaller}

The regression formulation of DiD:

$$Y_{it} = \beta_0 + \beta_1 \cdot \text{Treat}_i + \beta_2 \cdot \text{Post}_t + \beta_3 \cdot (\text{Treat}_i \times \text{Post}_t) + \varepsilon_{it}$$

:::: columns
::: {.column width="50%"}

| Parameter | Interpretation |
|-----------|---------------|
| $\beta_0$ | Mean outcome for control group, pre-period |
| $\beta_1$ | Pre-existing difference between groups |
| $\beta_2$ | Time trend (common to both groups) |
| $\beta_3$ | **DiD estimate** — the causal effect |

:::

::: {.column width="50%"}

|  | Pre | Post | $\Delta$ |
|--|-----|------|----------|
| **Control** | $\beta_0$ | $\beta_0 + \beta_2$ | $\beta_2$ |
| **Treated** | $\beta_0 + \beta_1$ | $\beta_0 + \beta_1 + \beta_2 + \beta_3$ | $\beta_2 + \beta_3$ |
| **Difference** | $\beta_1$ | $\beta_1 + \beta_3$ | $\beta_3$ |

The <mark>double difference</mark> removes both the time trend and the pre-existing group difference.

:::
::::

## Simulating a DiD {.smaller}

Let's simulate a policy that reduces substance use by <mark>3 percentage points</mark> in the treated state:

```{r}
#| echo: true

set.seed(2026)
n_per_cell <- 500
true_did_effect <- -3

did_data <- tibble(
    state = rep(c("Treated State", "Control State"), each = n_per_cell * 2),
    period = rep(rep(c("Pre", "Post"), each = n_per_cell), 2),
    treat = ifelse(state == "Treated State", 1, 0),
    post = ifelse(period == "Post", 1, 0)
) |>
    mutate(
        y = 25 +
            5 * treat +
            (-2) * post +
            true_did_effect * treat * post +
            rnorm(n(), 0, 8)
    )
```

- Both states have a baseline trend (decreasing by 2 over time)
- The treated state starts 5 points higher
- The treatment effect is $-3$ (3-point reduction in the treated state, post-period)

## Visualizing the DiD {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

did_means <- did_data |>
    summarise(mean_y = mean(y), .by = c(state, period)) |>
    mutate(period = factor(period, levels = c("Pre", "Post")))

ggplot(did_means, aes(x = period, y = mean_y, color = state, group = state)) +
    geom_point(size = 4) +
    geom_line(linewidth = 1.2) +
    geom_segment(
        data = did_means |>
            filter(state == "Treated State", period == "Post"),
        aes(
            x = "Post",
            xend = "Post",
            y = mean_y,
            yend = mean_y - true_did_effect
        ),
        linetype = "dashed",
        color = "grey50",
        linewidth = 0.8,
        arrow = arrow(length = unit(0.15, "cm"), ends = "both")
    ) +
    annotate(
        "text",
        x = 2.25,
        y = did_means$mean_y[
            did_means$state == "Treated State" &
                did_means$period == "Post"
        ] -
            true_did_effect / 2,
        label = paste0("DiD = ", true_did_effect),
        color = "grey30",
        size = 5,
        fontface = "bold"
    ) +
    scale_color_manual(
        values = c("Control State" = "#005030", "Treated State" = "#F47321")
    ) +
    labs(
        x = "Period",
        y = "Substance Use Prevalence (%)",
        color = "Group",
        title = "Difference-in-Differences Design",
        subtitle = "The dashed arrow shows the treatment effect (DiD estimand)"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "top")
```

## Estimating the DiD {.smaller}

```{r}
#| echo: true

did_model <- lm(y ~ treat + post + treat:post, data = did_data)

tidy(did_model, conf.int = TRUE) |>
    mutate(across(where(is.numeric), \(x) round(x, 2))) |>
    kable()
```

. . .

:::: columns
::: {.column width="50%"}

The interaction term `treat:post` is our <mark>DiD estimate</mark>:

- **Estimate**: close to the true effect of $-3$
- **Interpretation**: The treated state experienced an additional `r round(coef(did_model)["treat:post"], 1)`-point change in substance use compared to the control state

:::

::: {.column width="50%"}

::: {.callout-note}
## Reading the Other Coefficients

- `treat`: the pre-existing difference between states ($\approx 5$)
- `post`: the common time trend ($\approx -2$)
- `treat:post`: the causal effect, net of the trend and the baseline difference
:::

:::
::::

## Checking the Parallel Trends Assumption {.smaller}

:::: columns
::: {.column width="50%"}

The parallel trends assumption cannot be tested directly, but we can check whether it held <mark>before</mark> the intervention using multiple pre-treatment periods:

```{r}
#| echo: false
#| fig-width: 5
#| fig-height: 4

set.seed(2026)
multi_period <- expand_grid(
    state = c("Treated", "Control"),
    year = 2013:2020,
    id = 1:200
) |>
    mutate(
        treat = ifelse(state == "Treated", 1, 0),
        post = ifelse(year >= 2018, 1, 0),
        y = 30 +
            4 * treat +
            (-1.5) * (year - 2013) +
            (-3) * treat * post +
            rnorm(n(), 0, 6)
    )

multi_means <- multi_period |>
    summarise(mean_y = mean(y), .by = c(state, year))

ggplot(
    multi_means,
    aes(
        x = year,
        y = mean_y,
        color = state,
        group = state
    )
) +
    geom_point(size = 3) +
    geom_line(linewidth = 1) +
    geom_vline(
        xintercept = 2017.5,
        linetype = "dashed",
        color = "grey50"
    ) +
    annotate(
        "text",
        x = 2017.3,
        y = 32,
        label = "Intervention",
        angle = 90,
        vjust = -0.5,
        color = "grey40",
        size = 3.5
    ) +
    scale_color_manual(
        values = c(
            "Control" = "#005030",
            "Treated" = "#F47321"
        )
    ) +
    labs(
        x = "Year",
        y = "Outcome",
        color = "Group",
        title = "Pre-Trends Check"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top")
```

:::

::: {.column width="50%"}

<mark>What to look for:</mark>

- **Parallel pre-trends**: Before the intervention (left of dashed line), both groups follow the same trajectory
- **Divergence after**: After the intervention, the treated group deviates from the parallel path

::: {.callout-warning}
## When Parallel Trends Fails

If pre-treatment trends diverge, DiD is biased. Possible remedies:

1. **Add covariates** that explain the differential trend
2. **Use a different control group** with more similar trends
3. **Synthetic control methods** construct a weighted combination of control units that matches the treated unit's pre-trend
4. **Report sensitivity analyses** showing how much violation is needed to change conclusions
:::

:::
::::

## DiD: Practical Considerations {.smaller}

:::: columns
::: {.column width="50%"}

**Strengths of DiD:**

- Controls for all <mark>time-invariant</mark> unobserved confounders (anything constant within each group)
- Controls for common <mark>time trends</mark> (anything affecting both groups equally)
- Intuitive and transparent
- Widely used in policy evaluation

**When to use DiD:**

- A policy or program is implemented in some units but not others
- You have data from before and after the intervention
- You can identify a plausible comparison group
- Pre-trends are approximately parallel

:::

::: {.column width="50%"}

**Limitations:**

- <mark>Does not</mark> control for time-varying confounders that differ between groups
- Parallel trends is untestable for the post-period
- Standard errors need clustering at the group level when treatment is assigned at the group level

::: {.callout-tip}
## Extensions

Modern DiD methods handle more complex settings:

- **Staggered adoption**: different units adopt treatment at different times (`did` package, Callaway & Sant'Anna, 2021)
- **Multiple periods**: event-study designs that show dynamic effects
- **Continuous treatment**: generalizations beyond binary treatment
:::

:::
::::

# Part V: Putting It All Together {background-color="#005030"}

## Choosing a Method {.smaller}

:::: columns
::: {.column width="50%"}

| Method | Key Assumption | Best For |
|--------|---------------|----------|
| **Propensity Score (IPW/Matching)** | No unmeasured confounding | Cross-sectional observational data with rich covariates |
| **Difference-in-Differences** | Parallel trends | Panel data with a natural experiment or policy change |
| **Regression Discontinuity** | Continuity at cutoff | Assignment based on a running variable with a threshold |
| **Instrumental Variables** | Valid instrument | When unmeasured confounding exists but you have an exogenous source of variation |

:::

::: {.column width="50%"}

::: {.callout-important}
## No Method Is a Magic Bullet

Every causal inference method relies on <mark>assumptions</mark> that cannot be fully verified from the data. The strength of a causal argument comes from:

1. **Theoretical justification** for the assumptions
2. **Empirical checks** (balance, pre-trends, placebo tests)
3. **Sensitivity analyses** (how much violation is needed to change the conclusion?)
4. **Triangulation** across multiple methods and data sources
:::

:::
::::

## A Complete Simulation: Bias, IPW, and DiD {.smaller}

Let's compare all approaches on our simulated data:

```{r}
#| echo: true

results <- tibble(
    Method = c(
        "1. Naive (no adjustment)",
        "2. Adjust for confounder",
        "3. IPW",
        "4. RCT (gold standard)"
    ),
    Estimate = c(
        coef(lm(y_obs ~ treatment_obs, data = sim_data))["treatment_obs"],
        coef(lm(y_obs ~ treatment_obs + baseline_risk, data = sim_data))[
            "treatment_obs"
        ],
        coef(lm(y_obs ~ treatment_obs, data = sim_data, weights = ipw))[
            "treatment_obs"
        ],
        coef(lm(y_rct ~ treatment_rct, data = sim_data))["treatment_rct"]
    ),
    True_Effect = true_effect
) |>
    mutate(
        Bias = round(Estimate - True_Effect, 2),
        Estimate = round(Estimate, 2)
    )

kable(results)
```

. . .

> Methods 2–4 all recover estimates close to the truth. The naive estimate is biased. This illustrates the value of proper causal inference design.

## Key Takeaways {.smaller}

1. <mark>Causation requires more than association</mark>. The potential outcomes framework defines causation precisely through counterfactuals.

2. <mark>Randomization solves the problem</mark> by making treatment independent of all confounders. But randomization isn't always feasible.

3. <mark>Propensity score methods</mark> (IPW, matching) reconstruct balance from observational data by modeling the treatment assignment mechanism. They rely on **no unmeasured confounding**.

4. <mark>Difference-in-Differences</mark> leverages natural experiments and controls for time-invariant confounders by comparing changes over time. It relies on **parallel trends**.

5. <mark>No method is assumption-free</mark>. The credibility of causal claims rests on the plausibility of the identifying assumptions, empirical checks, and sensitivity analyses.

6. <mark>Design before analysis</mark>. Following Rubin's framework, construct balanced groups (or identify natural experiments) before examining outcomes.

## References {.smaller}

- Rubin, D. B. (2007). The design versus the analysis of observational studies for causal effects: Parallels with the design of randomized trials. *Statistics in Medicine, 26*(1), 20–36.
- Rosenbaum, P. R., & Rubin, D. B. (1983). The central role of the propensity score in observational studies for causal effects. *Biometrika, 70*(1), 41–55.
- Shadish, W. R. (2010). Campbell and Rubin: A primer and comparison of their approaches to causal inference in field settings. *Psychological Methods, 15*(1), 3–17.
- Callaway, B., & Sant'Anna, P. H. C. (2021). Difference-in-differences with multiple time periods. *Journal of Econometrics, 225*(2), 200–230.
- Imbens, G. W., & Rubin, D. B. (2015). *Causal Inference for Statistics, Social, and Biomedical Sciences*. Cambridge University Press.

## Questions?
